### 2016-07-23 Projects/DelCo_2012/Report/DelCo_2012.tex
initial edits to discussion section CAFS

This study focused on two key types of fire experiments; gas cooling and fire suppression. In each type of experiment a comparison was conducted between plain water and compressed air foam being discharged from nozzles of similar type with similar mass flow rates. In both types of experiments there were no significant measurable differences between the use of plain water or compressed air foam. These results were not surprising to some of the technical panel members. However other members, that use CAF very successfully in their fire departments every day, had additional questions. The research team worked to address those follow-up questions by conducting additional fire suppression experiments.  One set of experiments used only wood based fuels, to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. A final experimental series, used a larger fire room, with more ventilation.  In this final series, another change was made, the nozzle was hand held and could be moved as needed to effectively extinguish the fire. This section will discuss the results and provide potential explanations to support them.

@@

While CAF has been readily accepted for use in the wildland and wildland urban interface, the effectivenss of CAF for interior structural firefighting varies widely.  For example, does CAF provide improved gas cooling when compared to plan water?  During the planning of the experiments, some technical panel members thought that the excellent heat absorbing capacity of plain water would exceed the heat absorbing capabilities of CAFS.  Their position was based on the thought that CAF has injected air.  Since air is a great insulator, it slow heat transfer and result in less cooling. It was their belief that the reduced gas cooling capabilities of CAF may have led to firefighter injuries during interior fire attacks. Other member of the technical panel thought that the bubble structure would provide the same mass flowrate of water but with more surface area, which would enable a faster heat transfer rate and therefore, improve the cooling of the hot gas layer. There were also thoughts that the CAFS would coat the ceiling and eliminate energy feedback from the ceiling to the hot gas layer, thereby providing enhanced cooling.

The gas cooling experiments were designed to measure the ability of water streams and CAF streams to cool a hot gas layer, without any application of agent to the source of the heat.  The 3.35~m (11~ft) tall ceiling in the burn building allowed a hot gas layer,at least 1.83~m (6 ft) to form. During the testing, a minimum temperature of 250~$^{\circ}$C (482~$^{\circ}$F) at positions 1.83~m (6~ft) below the ceiling was attained prior to the beginning of another gas cooling experiment.

In each gas cooling experiment, the same mass flow rate of water was used for both the plain water stream and the CAF stream. Typically the volumetric flow rate of water was 7.57~L/s (120~gpm). The density of water is 1000~kg/m$^3$ (8.3~lb/gal), therefore the mass flow rate was 7.57~kg/s (16.7~lb/s). The average time of flow was approximately 15~s, therefore a total of 114~kg (251~lb) of water was introduced into each gas cooling experiment.

When a CAF stream was used it contained 0.3\% foam concentration by volume, the density of the foam concentrate is 1055~kg/m$^3$ (8.8~lb/gal), for a 15~s flow time, 0.34~kg (0.75~lb) of concentrate was used. In addition to the foam concentrate and water, the CAF stream has compressed air. The volume of air injected into the foam solution is 0.03~m$^3$/s (60~cfm). The density of the compressed air in the range of 100 to 150~psig, would be between 8.9~kg/m$^3$ (0.6 lb/ft$^3$) and 12.8~kg/m$^3$ (0.8 lb/ft$^3$). So a 15~s CAF flow would contain approximately 4.0~kg (8.8 lb) to 9.5~kg (21.0 lb) of air. Comparing the mass flows, and using the air mass based on the higher pressure, water still makes up more than 90~\% of the mass of the CAF stream.

The estimated total expanded uncertainty of the temperature measurements in these experiments is $\pm$ 15~\%.  In other words, for a difference between one measurement and another measurement to be considered significantly different, the difference would have to be greater by more than $\pm$ 15~\%.  If the difference between two measurements was $\pm$ 15~\% or less, the results are considered to be similar. That said, the estimated uncertainty is based on previous studies where suppression with a water based agent was included.

As pointed out in the results section, the conditions at TC Array 3 represented the ``'best'' gas cooling results with the least impact of agents hitting and cooling the thermocouples themselves. Under the given conditions of a residential scale room and the amount of water introduced into the hot gas layer, it should be no surprise that the water is doing the work. Air filled bubbles in the CAFS stream had no discernible impact, either positive or negative, in the gas cooling experiments.  At the same time there was no measurable negative effect of the air bubbles with regard to gas cooling.

The data from the thermocouples in TC Array 2, provide a sense of the impact of a combination of surface cooling and gas cooling.  For example, when the nozzle was in the "middle position" the straight or solid hose streams would impact near the middle of the ceiling of the room and they would create a fan pattern across the ceiling that would cause the majority of the water to impact the location of TC array 2 and the adjacent walls.  In this location, the data tended to favor water.  But again this did not represent gas cooling, this represented water or CAF hitting the thermocouples and cooling the surrounding gases. As a result the temperature reductions of the thermocouples in TC Array 2 are higher than those of the thermocouples in TC Array 3, as shown in \ref{fig:CAFS_Water_A2_mid}, and \ref{fig:CAFS_Water_A3_mid}.

TC Array 1 had the worse data fit or the widest spread of temperature change because sometimes, some of the thermcouples in the array were getting impacted by water or CAF while during other nozzle streams they were not. Examining the spray density figures in Appendix A.1, provides a range water collected in the bucket where TC Array 1 was located.  In many cases such as CAF Straight Stream Mid, the foam solution collected was near zero. In other cases such as CAF Fog Mid, the amount of foam solution colected was more than 3~kgs.  Similar variations occurred with water.  As a result, TC Array 1 did not provide consistant results for gas cooling.

@@

\subsection{Iowa State University Fire Flow Formula (1963)}
The Iowa Formula for determining the required fire flowrate for a given structure was developed, by Keith Royer and Bill Nelson, based on theoretical heat absorption capabilities of water. The required fire flowrate in gpm is equal to the volume of the fire compartment (in cubic feet) divided by 100:
\begin{equation}\label{eq:isu_form}
required fire flowrate = Volume / 100
\end{equation}
Due to inefficiencies in applying the water, some feel that the fire flow value from the Iowa State University formula should be multiplied by a factor of two to four~\cite{NFPA}. Applying the  based on floor area yields an application density of 3.3~Lpm/m$^2$ (0.08~gpm/ft$^2$) for ceiling heights of 2.44~m (8~ft) or 4.1~Lpm/m$^2$ (0.10~gpm/ft$^2$) for ceiling heights of 3.05~m (10~ft). Our fire rooms ranged in floor area from approximately 66~m$^3$ (2330~ft$^3$) to 125~m$^3$ (4410~ft$^3$). Therefore the required flowrate for the fire room in the single story structure based on the Iowa State University formula would be approximately 87~Lpm (23~gpm). The required flowrate for the fire room in the lower level of the two story would be approximately 168~Lpm (44~gpm).  While the formula has been widely taught over the years, there is some uncertainty about the application of the formula.  In some texts it mentions that the Iowa Formula provides the amount of water that needs to be applied in 30 seconds.  If that is the case then the flowrates given "per minute" must be doubled.  So in our case the required flowrates would increase to 174~Lpm (46~gpm)
and 336~Lpm (88 gpm), respectively.  During the experiments the flowrate was approximately, 454~Lpm (120~gpm), which exceeds the required flowrate of the Iowa Formula.  So the fact that all of the fires were quickly controlled is consistant with this methodology.

\subsection{National Fire Academy (1981)}
The National Fire Academy (NFA) developed a formula based on a study of fire flows, which were successful in controlling a large number of working fires~\cite{Klaene:1}. This formula sets the required fire flow, in gpm, equal to the floor area (in square feet) divided by 3.
\begin{equation}\label{eq:nfa_form}
required fire flowrate = Area / 3
\end{equation}

The fire rooms in these experiments ranged in floor area from approximately 27~$m^2$ (291~$ft^2$) to 51.4~$m^2$ (553~$ft^2$). Applying the NFA formula based on floor area yields an required flowrate for the single story fire room of 369~Lpm (97~gpm). The required flowrate for the fire room in the lower level of the two story would be approximately 700~Lpm (184~gpm).  Comparing our hosestream flowrates to the NFA Formula requirements indicates that 454~Lpm (120~gpm) was sufficient for the smaller room fire in the single story but was below the reuired flowrate of approximately 700~Lpm (184~gpm) for the large fire compartment in the two story structure.  This is not consistant with the observations of a rapid fire knockdown with both water and CAF hose streams flowing 454~Lpm (120~gpm).

Due to recent research conducted by NIST and the UL Firefighter Safety Research Institute (FSRI), the fire service has a re-newed interest in controlling the air available to a ventilation limited compartment fire.  By limiting the air (oxygen) available for combustion is one means of limting the heat release rate.  If the heat release rate of a fire is reduced due to a lack of oxygen needed for combustion, then in theory the amount of water needed to cool and suppress the fire should also be reduced.

Both of the formulas listed above do not overtly account for the ventilation of the fire structure only the size of the fire structure.  Further the most recent formula was developed in 1981, based on the fire ground experience of fire chiefs.  The experience that was shared was likely from fire that occurred in the 1960's and the 1970's.  It has been demonstrated that the synthetic fuels which fill the homes of the 2000's burn faster and hotter (higher heat release rate) than the natural fuels that filled the homes in the 1940's and 1950's.  The change of materials in our homes began in the 1960's and continues through to present day.  However it was unlikely that this change was factored in to the NFA Formula.

Shortly after the NFA equation was issued, another Federal agency, The National Bureau of Standards (now NIST) released a finding about the relationship of oxygen to the heat release rate of a fire.  They found that for many fuels, approximately 13.1~MJ of heat were generated for every kilogram of oxygen comsumed during the combustion.  This relationship then led to simple algorithms that addressed the relationship of the ventilation available o the room, in other words how much oxygen can flow through an opening, with the theoretical heat release rate which assumes stoichiometric conditions or 100~\% effecient burning where all of the fuel and the oxygen is consumed.

All of the fire conditions generated during these experiments were ventilation limited or ventilation controlled fires. In other words the conditions in the compart were fuel rich.  If the ventilation was increased to allow more oxygen to be supplied to the fuel gases available in the fire compartment and more products of combustion were exhausted from the fire compartment, the heat release rate of the fire would have increased. There are simple formulas which may provide some insight here, although for full disclosure this simple algorithm has not been validated with the geometry that we used in these experiments.  The purpose here is to get a sense of the amount of heat that the hose streams need to overcome.  This algorithm, just like the NFA formula has never been validated for this application.

\begin{equation} \label{eq:HRRSto_item}
\dot{Q}_{stoichiometric} = 1500 Area \sqrt{Height} \\
\end{equation}

In this equation, the maximum heat release rate, in~kW, based on airflow through the area of the opening, in~m^2, represented by Area. The other input needed is the height of the opening in~m, shown as Height.

@@

In all of the suppression experiments, the fire was controlled with total flow of 227~L (60~gal) or less of agent. During the course of the suppression experiments the fuel packages were modified to incorporate more wood based fuels.  The thinking of some of the technical panel members was that the synthetic fuels, such as polyurethane foam, burned mostly at the surface and therefore they were easily cooled and suppressed.  The wood based fuels, develop a char layer and therefore the burning would be "deep seated", which might provide a more significant suppression challenge.  Since the Class A foam solution has a reduced surface tension, which could allow the solution to penetrate and cool the burning wood based fuels better, it was thought that with more exposed, burning wood surface, the better the opportunity for CAF to demnostrate a difference.  As shown in Chapter 2, the wood fuel load constisted of stacks of wood pallets, plywood lined walls, and an exposed wood "ceiling" that was constructed of wood "i-beams" and oriented strand board.  The experiments conducted with the wood fuel load did not demonstrate a significant difference between water and CAF.

@@

\Chapter{Future Research}

@@

\section{Spray density figures for burn building tests}

@@

\section{Spray density figures for experimental structure tests}

>>>>

This study focused on two key types of fire experiments; gas cooling and fire suppression. In each type of experiment a comparison was conducted between plain water and compressed air foam being discharged from nozzles of similar type with similar mass flow rates. In both types of experiments there were no significant measurable differences between the use of plain water or compressed air foam. To expand upon the 2012 and 2013 tests described in~\cite{Mitchell:1}, additional fire suppression experiments were conducted. One set of experiments used only wood based fuels, to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. A final experimental series, used a larger fire room, with more ventilation. In this final series, another change was made, the nozzle was hand held and could be moved as needed to effectively extinguish the fire. This section will discuss the overall results of the study and provide potential explanations to support them.

@@

While CAF has been readily accepted for use in the wildland and wildland urban interface, the effectiveness of CAF for interior structural firefighting varies widely. Questions remain regarding whether CAF provides improved gas cooling when compared to plain water. During the planning of the experiments, there were a range of hypothesis regarding the outcome. Some technical panel members thought that the excellent heat absorbing capacity of plain water would exceed the heat absorbing capabilities of CAF. Their position was based on the thought that CAF has injected air. Since air is a great insulator, it slows heat transfer and results in less cooling. It was their belief that the reduced gas cooling capabilities of CAF may have led to firefighter injuries during interior fire attacks. Other members of the technical panel thought that the bubble structure would provide the same mass flowrate of water but with more surface area. The increased surface area would enable faster heat transfer rate and therefore, improve the cooling of the hot gas layer. There were also thoughts that the CAFS would coat the ceiling and eliminate energy feedback from the ceiling to the hot gas layer, thereby providing enhanced cooling.

The gas cooling experiments were designed to measure the ability of water streams and CAF streams to cool a hot gas layer, without any application of agent to the source of the heat. The 3.35~m (11~ft) tall ceiling in the burn building allowed a hot gas layer, at least 1.83~m (6 ft) to form. During the testing, a minimum temperature of 250~$^{\circ}$C (482~$^{\circ}$F) at positions 1.83~m (6~ft) below the ceiling was attained prior to the beginning of another gas cooling experiment. In each gas cooling experiment, the same mass flow rate of water was used for both the plain water stream and the CAF stream. Typically the volumetric flow rate of water was 7.57~L/s (120~gpm). The density of water is 1000~kg/m$^3$ (8.3~lb/gal), therefore the mass flow rate was 7.57~kg/s (16.7~lb/s). The average time of flow was approximately 15~s, therefore a total of 114~kg (251~lb) of water was introduced into each gas cooling experiment.

When a CAF stream was used it contained 0.3~\% foam concentration by volume with the density of the foam concentrate is 1055~kg/m$^3$ (8.8~lb/gal). For a 15~s flow time, 0.34~kg (0.75~lb) of concentrate was used. In addition to the foam concentrate and water, the CAF stream has compressed air. The volume of air injected into the foam solution is 0.03~m$^3$/s (60~cfm). The density of the compressed air is in the range of 100 to 150~psig, which would be between 8.9~kg/m$^3$ (0.6~lb/ft$^3$) and 12.8~kg/m$^3$ (0.8~lb/ft$^3$). Therefore, a 15~s CAF flow would contain approximately 4.0~kg (8.8~lb) to 9.5~kg (21.0~lb) of air. Comparing the mass flows, and using the air mass based on the higher pressure, water still makes up more than 90~\% of the mass of the CAF stream.

The estimated total expanded uncertainty of the temperature measurements in these experiments is $\pm$ 15~\% (see Section~\ref{subsec:Uncertainty}). In other words, for a difference between one measurement and another measurement to be considered significantly different, the difference would have to be greater by more than $\pm$ 15~\%. If the difference between two measurements was $\pm$ 15~\% or less, the results are considered to be similar. That said, the estimated uncertainty is based on previous studies where suppression with a water based agent was included.

As pointed out in the results section, the conditions at TC Array 3 represented the ``'best'' gas cooling results because that position had the least impact of agents hitting and cooling the thermocouples themselves. Under the given conditions of a residential scale room and the amount of water introduced into the hot gas layer, it should be no surprise that the water is doing the work. Air filled bubbles in the CAF stream had no discernible impact, either positive or negative, in the gas cooling experiments. At the same time there was no measurable negative effect of the air bubbles with regard to gas cooling.

The data from the thermocouples in TC Array 2, provide a sense of the impact of a combination of surface cooling and gas cooling. For example, when the nozzle was in the ``middle position'' the straight or solid hose streams would impact near the middle of the ceiling of the room and they would create a fan pattern across the ceiling that would cause the majority of the water to impact the location of TC Array 2 and the adjacent walls. In this location, the data tended to favor water. But again this did not represent gas cooling, this represented some of the water or CAF hitting the thermocouples and cooling the surrounding gases. As a result, temperature reductions of the thermocouples in TC Array 2 are higher than those of the thermocouples in TC Array 3, as shown in \ref{fig:CAFS_Water_A2_mid}, and \ref{fig:CAFS_Water_A3_mid} respectively.

TC Array 1 had the widest spread of temperature change. Depending on the nozzle configuration, there were tests where the Array 1 thermocouples were directly impacted by water or CAF and other tests where there was no impact. Examination of the spray density figures in Appendix~\ref{spray_bb}, provides the range of agent collected in the bucket where TC Array 1 was located. In many cases such as CAF Straight Stream Mid, the foam solution collected was near zero. In other cases such as CAF Fog Mid, the amount of foam solution collected was more than 3~kgs. Similar variations occurred with water. As a result, TC Array 1 did not provide consistent results for gas cooling.

@@

\subsection{Iowa State University Fire Flow Formula (1963)}
The Iowa Formula for determining the required fire flowrate for a given structure was developed, by Keith Royer and Bill Nelson, based on theoretical heat absorption capabilities of water. The required fire flowrate, ($\dot{V}$), in gpm is equal to the volume ($V$) of the fire compartment (in cubic feet) divided by 100:
\begin{equation}\label{eq:isu_form}
\dot{V} = V / 100
\end{equation}
Due to inefficiencies in applying the water, some feel that the fire flow value from the Iowa State University formula should be multiplied by a factor of two to four~\cite{NFPA}. Applying the formula based on floor area yields an application density of 3.3~Lpm/m$^2$ (0.08~gpm/ft$^2$) for ceiling heights of 2.44~m (8~ft) or 4.1~Lpm/m$^2$ (0.10~gpm/ft$^2$) for ceiling heights of 3.05~m (10~ft). The fire rooms in these experiments ranged in floor area from approximately 66~m$^3$ (2330~ft$^3$) to 125~m$^3$ (4410~ft$^3$). Therefore the required flowrate for the fire room in the single story structure based on the Iowa State University formula would be approximately 87~Lpm (23~gpm). The required flowrate for the fire room in the lower level of the two story structure would be approximately 168~Lpm (44~gpm). While the formula has been widely taught over the years, there is some uncertainty about the application of the formula. Some texts mention that the Iowa State Formula provides the amount of water that needs to be applied in 30 seconds. If that is the case, then the flowrates given ``per minute'' must be doubled. The required flowrates would increase to 174~Lpm (46~gpm) and 336~Lpm (88~gpm), respectively. During the experiments the flowrate was approximately, 454~Lpm (120~gpm), which exceeds the required flowrate of the Iowa State Formula. The fact that all of the fires were quickly controlled  during the experiments is consistent with this methodology.

\subsection{National Fire Academy (1981)}
The National Fire Academy (NFA) developed a formula based on a study of fire flows, which were successful in controlling a large number of working fires~\cite{Klaene:1}. This formula sets the required fire flow, ($\dot{V}$), in gpm equal to the floor area ($A$) (in square feet) divided by 3.
\begin{equation}\label{eq:nfa_form}
\dot{V} = A / 3
\end{equation}

The fire rooms in these experiments ranged in floor area from approximately 27~$m^2$ (291~$ft^2$) to 51.4~$m^2$ (553~$ft^2$). Applying the NFA formula based on floor area yields a required flowrate for the single story fire room of 369~Lpm (97~gpm). The required flowrate for the fire room in the lower level of the two story would be approximately 700~Lpm (184~gpm). Comparing the hose stream flowrates to the NFA Formula requirements indicates that 454~Lpm (120~gpm) was sufficient for the smaller room fire in the single story but was below the required flowrate of approximately 700~Lpm (184~gpm) for the large fire compartment in the two story structure. This is not consistent with the observations of a rapid fire knockdown with both water and CAF hose streams flowing 454~Lpm (120~gpm).

Due to recent research conducted by NIST and the UL Firefighter Safety Research Institute (FSRI), the fire service has a re-newed interest in controlling the air available to a ventilation limited compartment fire. Limiting the air (oxygen) available for combustion is one means of limiting the heat release rate. If the heat release rate of a fire is reduced due to a lack of oxygen needed for combustion, then in theory, the amount of water needed to cool and suppress the fire should also be reduced.

Both of the formulas listed above do not overtly account for the ventilation of the fire structure, only the size of the fire structure. Further the most recent formula was developed in 1981, based on the fire ground experience of fire chiefs. The experience that was shared was likely from fires that occurred in the 1960's and the 1970's. It has been demonstrated that the synthetic fuels which fill the homes of the 2000's burn faster and with higher heat release rates than the natural fuels that filled the homes in the 1940's and 1950's. The change of materials in homes began in the 1960's and continues through to present day. However, it was unlikely that this change was factored in to the NFA Formula.

Shortly after the NFA equation was issued, another Federal agency, The National Bureau of Standards (now NIST) released a finding about the relationship of oxygen to the heat release rate of a fire. NIST found that for many fuels, approximately 13.1~MJ of heat are generated for every kilogram of oxygen consumed during the combustion. This relationship then led to simple algorithms that addressed the how much oxygen can flow through an opening with a theoretical heat release rate that assumes stoichiometric conditions (100~\% efficient burning where all of the fuel and the oxygen is consumed).

All of the fire conditions generated during these experiments were ventilation limited or ventilation controlled fires. Essentially, the conditions in the compartment were fuel rich. If the ventilation was increased to allow more oxygen to be supplied to the fuel gases available in the fire compartment and more products of combustion were exhausted from the fire compartment, the heat release rate of the fire would have increased. There are simple formulas which may provide some insight here, although for full disclosure this simple algorithm has not been validated with the geometry that we used in these experiments.  The purpose here is to get a sense of the amount of heat that the hose streams need to overcome.  This algorithm, just like the NFA formula has never been validated for this application.

\begin{equation} \label{eq:HRRSto_item}
\dot{Q}_{stoichiometric} = 1500 A_0 \sqrt{H_0} \\
\end{equation}

In this equation, the maximum heat release rate, in~kW, based on airflow through the area of the opening, in m$^2$, represented by Area. The other input needed is the height of the opening in m, shown as Height.

@@

In all of the suppression experiments, the fire was controlled with total flow of 227~L (60~gal) or less of agent. During the course of the suppression experiments the fuel packages were modified to incorporate more wood based fuels.  The thinking of some of the technical panel members was that the synthetic fuels, such as polyurethane foam, burned mostly at the surface and therefore they were easily cooled and suppressed.  The wood based fuels, develop a char layer and therefore the burning would be ``deep seated'', which might provide a more significant suppression challenge.  Since the Class A foam solution has a reduced surface tension, which could allow the solution to penetrate and cool the burning wood based fuels better, it was thought that with more exposed, burning wood surface, the better the opportunity for CAF to demnostrate a difference.  As shown in Chapter 2, the wood fuel load constisted of stacks of wood pallets, plywood lined walls, and an exposed wood ``ceiling'' that was constructed of wood ``i-beams'' and oriented strand board.  The experiments conducted with the wood fuel load did not demonstrate a significant difference between water and CAF.

@@

\section{Future Research}

@@

\section{Spray density figures for burn building tests}
\label{app:spray_bb}

@@

\section{Spray density figures for experimental structure tests}
\label{app:spray_es}
